---
title: "Killer whale Clock"
author: "Caila Kucheravy"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

Run the killer whale skin clock developed by Parsons et al. (2023). 

Prep: 
```{r, message = FALSE, results = 'hide'}
setwd("~/Documents/Master's/Analysis/Epigenetic Aging/Killer Whales")

library(tidyverse)
library(glmnet)
```

Load updated sample sheet:
```{r}
sample_sheet <- readRDS('output/updated_sample_sheet_combined_KillerWhale_array.rds') 
```

Load the killer whale skin clock, select the correct columns and filter out NAs to have only CpGs used in the clock: 
```{r, message = FALSE, results = 'hide'}
clock <- read_csv('input/Table.WhaleS3.SkinClockCoef.csv')

kw_clock <- clock %>% 
  select(var, Coef.Killerwhale.Skin.Sqrt) %>%      # Select the correct column for the KW clock 
  filter(!is.na(Coef.Killerwhale.Skin.Sqrt))       # Filter out CpGs not used for this clock (50 CpGs)
```

Load normalized beta values: 
```{r}
kw_betas <- readRDS('output/tbetas_corrected_combined_KillerWhale_array.rds')

# Filter for CpGs used in clock - 50 for killer whales
kw_betas_filtered <- kw_betas %>% 
  select(any_of(kw_clock$var))
```

The age transformation used in the paper is: sqrt(Age+1)=Age. Need to back-transform for final age.

Form a weighted linear combination of the CpGs for killer whales: 
```{r}
# Pivot clock wider to match beta table 
kw_clock_wide <- kw_clock %>% 
  pivot_wider(names_from = var, values_from = Coef.Killerwhale.Skin.Sqrt)

# Multiply beta values by the clock weights:
est_ages_kw <- data.frame(mapply('*', kw_betas_filtered, kw_clock_wide[,2:51])) 

# Sum values, add intercept
est_ages_kw <- est_ages_kw %>% 
  mutate(Sum = rowSums(est_ages_kw)) %>% 
  mutate(Intercept = kw_clock_wide$`(Intercept)`) 

# Sum intercept and weighted beta values
est_ages_kw <- est_ages_kw %>%
  mutate(Ages = rowSums(est_ages_kw[,c("Sum", "Intercept")]))

# Age transformation: DNAmAge = F^(-1) (x*beta)
est_ages_kw <- est_ages_kw %>%
  mutate(Age_Transformed = (est_ages_kw$Ages^2) - 1)

est_ages_kw$Age_Transformed
```

Add ages to sample sheet:
```{r}
# Add column with basename back to dataframe with ages
est_ages_kw$chip.ID.loc <- sample_sheet$chip.ID.loc    

# Select chip ID and ages to join with sample sheet
DNAm_ages_kw <- est_ages_kw %>%                         
  select(chip.ID.loc, Age_Transformed)

# Join ages with sample sheet
kw_ages <- sample_sheet %>%             
  left_join(DNAm_ages_kw, by = "chip.ID.loc")
```

Take a look at the duplicates:
```{r}
# Note that KW-2019-06 was a technical replicate (from the same DNA sample), while the others were from different skin samples of the same whale. 
duplicates <- kw_ages %>% 
  arrange(block) %>% 
  group_by(block) %>%
  filter(n() > 1)
duplicates
```

Most of the ages estimate for duplicate samples are quite close, but some of the recaptures don't reflect the difference in age between captures.

We also have a few known-age individuals. Unfortunately, two of the known-age individuals (ARSQ-xx-1379, 34yo and KW-CH-2011, 35 yo) had to be removed in the quality control stage.

We can compare the DNAm estimated age to the known age (determined from GLGs):
```{r}
known_ages <- data.frame(
  sampleId = c("ARRB-xx-1291", "ARSQ-xx-1397", "KW-2022-PI-01"),
  Known_Age = c(28, 6, "In progress")
)

est_ages <- kw_ages %>% 
  select(sampleId, Age_Transformed) %>% 
  filter(sampleId %in% c("ARRB-xx-1291", "ARSQ-xx-1397", "KW-2022-PI-01"))

compare_known_ages <- known_ages %>% 
  left_join(est_ages, by = "sampleId") 
compare_known_ages
```

Remove the duplicates and add the difference between year and 2022:
```{r}
# Taking the first age for now - but might want to take average or something for final ages.
kw_ages_dupsRemoved <- kw_ages %>% 
  mutate(duplicate = duplicated(block)) %>% 
  filter(!duplicate == "TRUE")

# Add the difference in years so that the age structure represents what it would be in 2022:
kw_ages_dupsRemoved <- kw_ages_dupsRemoved %>% 
  select(!duplicate) %>% 
  mutate(diffYear = 2022 - kw_ages_dupsRemoved$Year) %>% 
  mutate(adj_age = Age_Transformed + diffYear)

#Write the age file to csv: 
#write.csv(kw_ages_dupsRemoved, "kw_ages.csv")

# Keep only Cumberland Sound & Northern Baffin Island samples ("High Arctic" group):
Locations <- c("Cumberland Sound", "Eclipse Sound", "Newfoundland", "Saint Pierre et Miquelon")

kw_ages_HA <- kw_ages_dupsRemoved %>% 
  filter(Location %in% Locations) %>% 
  # And the one 2013 CS sample that grouped with the Greenland samples
  filter(!sampleId == "ARPG-2013-01")
```

Plot the data (unadjusted age):
```{r}
cols <- c("tomato3", "steelblue3")

ggplot(kw_ages_dupsRemoved, aes(x = Age_Transformed, fill = Sex)) + 
  geom_histogram() +
  xlab("DNAm Age") +
  #scale_x_continuous("DNAm Age", limits = c(0,55), breaks = c(0,5,10,15,20,25,30,35,40,45,50,55)) + 
  scale_fill_manual(values = cols, 
                      labels = c("Female", "Male")) + 
  theme_bw() + 
  theme(axis.text = element_text(size=14),
        axis.title = element_text(size=18),
        legend.title = element_blank(),
        legend.text = element_text(size = 14),
        legend.position = c(0.82,0.8),
        legend.background = element_blank())
```

```{r}
males <- kw_ages_dupsRemoved %>% 
  filter(Sex == "M")
females <- kw_ages_dupsRemoved %>% 
  filter(Sex == "F")

g <- ggplot() + 
  geom_histogram( data = females,
    aes(x = Age_Transformed, y = after_stat(count)),
    fill="tomato3") + 
  geom_histogram( data = males,
    aes(x = Age_Transformed, y = -after_stat(count)), 
    fill= "steelblue") + 
  theme_bw()

#tiff("Age_Distribution_2.tiff", units="in", width=8, height=5, res=400)

g + 
  coord_flip() + 
  xlab("DNAm Age") +
  #scale_x_continuous(limits = c(0,55), breaks = c(0,10,20,30,40,50)) + 
  #scale_y_continuous("DNAm Age", limits = c(-5,5), breaks = c(-5,-3,-1,1,3,5))
  theme(axis.text = element_text(size=14),
        axis.title = element_text(size=16))

#dev.off()
```

Plot the data (adjusted 2022 age):
```{r}
#tiff("age_structure_no.calves.tiff", units="in", width=6, height=4, res=500)

ggplot(kw_ages_HA, aes(x = adj_age, fill = Sex)) + 
  geom_histogram() +
  xlab("DNAm Age") +
  #scale_x_continuous("DNAm Age", limits = c(0,55), breaks = c(0,5,10,15,20,25,30,35,40,45,50,55)) + 
  scale_fill_manual(values = cols, 
                      labels = c("Female", "Male")) + 
  theme_bw() + 
  theme(axis.text = element_text(size=14),
        axis.title = element_text(size=18),
        legend.title = element_blank(),
        legend.text = element_text(size = 14),
        legend.position = c(0.82,0.8),
        legend.background = element_blank())

#dev.off()
```

Summarize data: 
```{r}
males <- kw_ages_dupsRemoved %>% 
  filter(Sex == "M")

females <- kw_ages_dupsRemoved %>% 
  filter(Sex == "F")

length(males$Age_Transformed)
summary(males$Age_Transformed)
summary(males$adj_age)

length(females$Age_Transformed)
summary(females$Age_Transformed)
summary(females$adj_age)

# Juveniles
kw_ages_dupsRemoved %>% 
  dplyr::count(adj_age < 10)
24/76

# Reproductive adults
kw_ages_dupsRemoved %>% 
  dplyr::count(adj_age < 35)
(75-24)/76

# Post-reproductive adults
kw_ages_dupsRemoved %>% 
  dplyr::count(adj_age > 35)
1/76
```



