---
title: "Killer whale Clock"
author: "Caila Kucheravy"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

Run the killer whale skin clock developed by Parsons et al. (2023). 

Prep: 
```{r, message = FALSE, results = 'hide'}
setwd("~/Documents/Master's/Analysis/Epigenetic Aging/Killer Whales")

library(tidyverse)
#library(glmnet)
```

Load updated sample sheet:
```{r}
sample_sheet <- readRDS('output/updated_sample_sheet_combined_KillerWhale_array.rds') 
```

Load the killer whale skin clock, select the correct columns and filter out NAs to have only CpGs used in the clock: 
```{r, message = FALSE, results = 'hide'}
clock <- read_csv('input/Table.WhaleS3.SkinClockCoef.csv')

kw_clock <- clock %>% 
  select(var, Coef.Killerwhale.Skin.Sqrt) %>%      # Select the correct column for the KW clock 
  filter(!is.na(Coef.Killerwhale.Skin.Sqrt))       # Filter out CpGs not used for this clock (50 CpGs)
```

Load normalized beta values: 
```{r}
kw_betas <- readRDS('output/tbetas_corrected_combined_KillerWhale_array_August2024_redo3.rds')

# Filter for CpGs used in clock - 50 for killer whales
kw_betas_filtered <- kw_betas %>% 
  select(any_of(kw_clock$var))
```

The age transformation used in the paper is: sqrt(Age+1)=Age. Need to back-transform for final age.

Form a weighted linear combination of the CpGs for killer whales: 
```{r}
# Pivot clock wider to match beta table 
kw_clock_wide <- kw_clock %>% 
  pivot_wider(names_from = var, values_from = Coef.Killerwhale.Skin.Sqrt)

# Multiply beta values by the clock weights:
est_ages_kw <- data.frame(mapply('*', kw_betas_filtered, kw_clock_wide[,2:51])) 

# Sum values, add intercept
est_ages_kw <- est_ages_kw %>% 
  mutate(Sum = rowSums(est_ages_kw)) %>% 
  mutate(Intercept = kw_clock_wide$`(Intercept)`) 

# Sum intercept and weighted beta values
est_ages_kw <- est_ages_kw %>%
  mutate(Ages = rowSums(est_ages_kw[,c("Sum", "Intercept")]))

# Age transformation: DNAmAge = F^(-1) (x*beta)
est_ages_kw <- est_ages_kw %>%
  mutate(Age_Transformed = (est_ages_kw$Ages^2) - 1)

est_ages_kw$Age_Transformed
```

Add ages to sample sheet:
```{r}
# Add column with basename back to dataframe with ages
est_ages_kw$chip.ID.loc <- sample_sheet$chip.ID.loc    

# Select chip ID and ages to join with sample sheet
DNAm_ages_kw <- est_ages_kw %>%                         
  select(chip.ID.loc, Age_Transformed)

# Join ages with sample sheet
kw_ages <- sample_sheet %>%             
  left_join(DNAm_ages_kw, by = "chip.ID.loc")

# Save csv file of all ages
# write.csv(kw_ages, "kw_ages_August2024")
```

Remove the "bad" samples from the epigenetic array: 
```{r}
bad <- c("KW-2021-PG-03", 
         "KW-2021-PG-04", 
         "KW-2021-PG-05", 
         "KW-2021-PG-06",
         "KW-2021-PG-08", 
         "KW-2021-PG-09", 
         "KW-2021-PG-11", 
         "KW-2021-PG-XX", 
         "KW-2019-01",
         "KW-2019-02",
         "GRNL-KW-2021-01",
         "GRNL-KW-2021-03",
         "OR21-1",
         "OR21-2",
         "KW-Nfld-22-25",
         "KW-2022-PI-01"
         )

kw_ages <- kw_ages  %>% 
  filter(!sampleId %in% bad)
```

Take a look at the duplicates:
```{r}
# Note that KW-2019-06 was a technical replicate (from the same DNA sample), while the others were from different skin samples of the same whale. 
duplicates <- kw_ages %>% 
  select(block, sampleId, Year, Location, Sex, Age_Transformed) %>% 
  arrange(block) %>% 
  group_by(block) %>%
  filter(n() > 1)
duplicates

#write.csv(duplicates, "duplicate_epigenetic_ages_August2024.csv")
```

Most of the ages estimate for duplicate samples are quite close, but some of the recaptures don't reflect the difference in age between captures.

Try plotting the differences in estimated ages across time.
```{r}
cols <- c("tomato3", "steelblue3")

# Plot difference in estimated ages across time
ggplot(duplicates, aes(x = Year, 
                       y = Age_Transformed, 
                       group = block,
                       #label = sampleId,
                       color = Sex)) + 
  geom_point() + 
  geom_line() + 
  #geom_text(hjust = -0.1, vjust = -0.1, cex = 2) +
  scale_x_continuous(limits = c(2013, 2020), breaks = seq(2013, 2020, by = 1)) +
  scale_y_continuous(breaks = seq(0, 30, by = 2)) +
  scale_color_manual(values = cols,
                     labels = c("Female", "Male")) +
  labs(x = "Year of Capture", y = "DNAm Age") +
  theme_bw() + 
  theme(axis.text = element_text(size=14),
        axis.title = element_text(size=18),
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        legend.position = c(0.15,0.79),
        legend.background = element_blank())

#ggsave("Plots/age_difference_by_year.png", width = 8, height = 6, dpi = 300)
```

We also have a few known-age individuals. Unfortunately, two of the known-age individuals (ARSQ-xx-1379, 34yo and KW-CH-2011, 35 yo) had to be removed in the quality control stage, and the 2022 sample was mixed in the great sample disaster.

We can compare the DNAm estimated age to the known age (determined from GLGs):
```{r}
known_ages <- data.frame(
  sampleId = c("ARRB-xx-1291", "ARSQ-xx-1397"),
  Known_Age = c(28, 6))

est_ages <- kw_ages %>% 
  select(sampleId, Age_Transformed) %>% 
  filter(sampleId %in% c("ARRB-xx-1291", "ARSQ-xx-1397"))

compare_known_ages <- known_ages %>% 
  left_join(est_ages, by = "sampleId") %>% 
  mutate(Difference = Age_Transformed - Known_Age)
compare_known_ages
```

Take the first of duplicate samples, then remove duplicates and individuals not in the "High Arctic" population.
```{r}
# Remove duplicates - take the first sample
kw_ages_dupsRemoved <- kw_ages %>%
  arrange(block) %>% 
  mutate(duplicate = duplicated(block)) %>% 
  filter(!duplicate == "TRUE")

# Keep only Northern Baffin Island locations since this is where the mark recapture is run
kw_ages_HA <- kw_ages_dupsRemoved  %>% 
  filter(Location %in% c("Cumberland Sound", "Eclipse Sound")) %>% 
  # And the one 2013 CS sample that grouped with the Greenland samples
  filter(!sampleId == "ARPG-2013-01")
```

Plot the data (unadjusted - age at year of capture):
```{r}
# Set plot colors 
#cols <- c("tomato3", "steelblue3")
cols <- c("#dc8374", "#83b4dc")

# Calculate means for males & females 
males <- kw_ages_HA %>% 
  filter(Sex == "M")
mean_males <- mean(males$Age_Transformed)
median_males <- median(males$Age_Transformed)

females <- kw_ages_HA %>% 
  filter(Sex == "F")
mean_females <- mean(females$Age_Transformed)
median_females <- median(females$Age_Transformed)

#tiff("Plots/Age_Distribution_August2024_2.tiff", units="in", width=8, height=5, res=400)

# Plot
ggplot(kw_ages_HA, aes(x = Age_Transformed, fill = Sex)) + 
  geom_histogram(position = "dodge") +
  scale_x_continuous("DNAm Age", limits = c(0,35), breaks = seq(0, 35, by = 2)) + 
  scale_y_continuous(limits = c(0,6), breaks = seq(0, 6, by = 2)) + 
  scale_fill_manual(values = cols, 
                    labels = c("Female", "Male")) + 
  geom_vline(xintercept = mean_males, col = "#37678f", lty = 1, size = 1) + 
  geom_vline(xintercept = mean_females, col = "#8f3727", lty = 1, size = 1) + 
  theme_classic() + 
  theme(axis.text = element_text(size=14),
        axis.title = element_text(size=18),
        legend.title = element_blank(),
        legend.text = element_text(size = 18),
        legend.position = c(0.82,0.8),
        legend.background = element_blank())
  
#dev.off()
```

To get what the age structure would be in a given year, adjust ages to year 2020:
```{r}
kw_ages_HA <- kw_ages_HA %>% 
  mutate(diff_Year = (2022 - Year)) %>% 
  mutate(adj_age = (Age_Transformed + diff_Year))

#write.csv(kw_ages_HA, "kw_ages_HA_August2024.csv")
```

Plot the adjusted data (age corrected to 2020):
```{r}
# Calculate means for males & females 
mean_adj_males <- mean(males$adj_age)
median_adj_males <- median(males$adj_age)

mean_adj_females <- mean(females$adj_age)
median_adj_females <- median(females$adj_age)

#tiff("Plots/Adj_Age_Distribution_August2024_2.tiff", units="in", width=8, height=5, res=400)

# Plot
ggplot(kw_ages_HA, aes(x = adj_age, fill = Sex)) + 
  geom_histogram(position = "dodge") +
  xlab("DNAm Age") +
  scale_x_continuous("Adjusted DNAm Age", limits = c(0,35), breaks = seq(0, 35, by = 2)) + 
  scale_y_continuous(limits = c(0,6), breaks = seq(0, 6, by = 2)) + 
  scale_fill_manual(values = cols, 
                      labels = c("Female", "Male")) + 
  geom_vline(xintercept = mean_adj_males, col = "#37678f", lty = 1, size = 1) + 
  #geom_vline(xintercept = median_males, col = "steelblue3", lty = 2) + 
  geom_vline(xintercept = mean_adj_females, col = "#8f3727", lty = 1, size = 1) + 
  #geom_vline(xintercept = median_females, col = "tomato3", lty = 2) + 
  theme_classic() + 
  theme(axis.text = element_text(size=14),
        axis.title = element_text(size=18),
        legend.title = element_blank(),
        legend.position = "none")

#dev.off()
```

Summarize data: 
```{r}
# Number of samples
length(kw_ages_HA$sampleId)

# Summary of unadjusted age 
summary(kw_ages_HA$Age_Transformed)

# Summary of adjusted age 
summary(kw_ages_HA$adj_age)

# Summary for males: 
males <- kw_ages_HA %>% 
  filter(Sex == "M")

length(males$sampleId)
summary(males$Age_Transformed)
summary(males$adj_age)

# Proportion of males
length(males$adj_age)/length(kw_ages_HA$adj_age)

# Summary for females
females <- kw_ages_HA %>% 
  filter(Sex == "F")

length(females$sampleId)
summary(females$Age_Transformed)
summary(females$adj_age)

# Proportion of females
length(females$adj_age)/length(kw_ages_HA$adj_age)

# Juveniles
kw_ages_HA %>% 
  dplyr::count(adj_age < 13)
26/51

# Reproductive adults
kw_ages_HA %>% 
  dplyr::count(adj_age < 35)
25/51

# Post-reproductive adults
kw_ages_HA %>% 
  dplyr::count(adj_age > 35)

```






```{r}
g <- ggplot() + 
  geom_histogram( data = females,
    aes(x = Age_Transformed, y = after_stat(count)),
    fill="#dc8374") + 
  geom_histogram( data = males,
    aes(x = Age_Transformed, y = -after_stat(count)), 
    fill= "#83b4dc") + 
  theme_bw()

#tiff("Age_Distribution_2.tiff", units="in", width=8, height=5, res=400)

age_plot <- g + 
  coord_flip() + 
  scale_x_continuous("DNAm Age", limits = c(0,35), breaks = seq(0, 35, by = 5)) +
  scale_y_continuous("Count", limits = c(-5,5), breaks = seq(-5,5, by = 1)) +
  geom_segment(aes(x = mean_males, xend = mean_males, y = 0, yend = -5), col = "#37678f", lty = 1, size = 1.3) + 
  geom_segment(aes(x = mean_females, xend = mean_females, y = 0, yend = 5), col = "#8f3727", lty = 1, size = 1.3) +
  #annotate(geom = "text", label = "Male", x = 0.5, y = -5, color = "#37678f", cex = 7, hjust = 0) + 
  theme(axis.text = element_text(size=14),
        axis.title = element_text(size=16)) 

#dev.off()
```

```{r}
g2 <- ggplot() + 
  geom_histogram( data = females,
    aes(x = adj_age, y = after_stat(count)),
    fill="#dc8374") + 
  geom_histogram( data = males,
    aes(x = adj_age, y = -after_stat(count)), 
    fill= "#83b4dc") + 
  theme_bw()

#tiff("Age_Distribution_2.tiff", units="in", width=8, height=5, res=400)

g2 + 
  coord_flip() + 
  scale_x_continuous("DNAm Age", limits = c(0,35), breaks = seq(0, 35, by = 5)) +
  scale_y_continuous("Count", limits = c(-6,6), breaks = seq(-6,6, by = 1)) +
  geom_segment(aes(x = mean_adj_males, xend = mean_adj_males, y = 0, yend = -6), col = "#37678f", lty = 1, size = 1.3) + 
  geom_segment(aes(x = mean_adj_females, xend = mean_adj_females, y = 0, yend = 6), col = "#8f3727", lty = 1, size = 1.3) +
  #annotate(geom = "text", label = "Male", x = 0.5, y = -5, color = "#37678f", cex = 7, hjust = 0) + 
  theme(axis.text = element_text(size=14),
        axis.title = element_text(size=16)) 

#dev.off()
```
