---
title: "2.3 Genetic Mark Recapture"
date: "`r Sys.Date()`"
output: pdf_document
---

Capture mark recapture of Arctic killer whales using genetic (whole genome) mark recapture. Using the same statistical methods as Kyle did for his estimate (Lefort et al. 2020) and in 2.1_PhotoID_CMR. 

### POPAN Jolly Seber 

Assumptions for Jolly-Seber Mark Recapture models: 
1. Animals retain their tags throughout the experiment 
2. Tags are read properly
3. Sampling is instantaneous
4. Survival probabilities are the same for all animals (marked and unmarked) between each pair of sampling occasions (homogenous survival)
5. Catchability is the same for all animals (marked and unmarked) at each sampling occasion (homogenous catchability)
6. The study area is constant

In the case of killer whales, do we meet the assumptions?
1. Yes - most nicks and scars used for ID are retained through the life of the animal.
2. Yes - we can assume that identified individuals are re-identified reliably. However, there are ways to account for identification error - to be explored later. 
3. Yes - sampling period is short (1 to a few days)
4. Probably? - while survival probabilities might differ between sex and age classes, being "marked" does not affect an individual's survival
5. Unlikely? - Equal catchability could be affected by: 
    - Behaviour: some individuals/groups may be more likely to approach the boat, and thus we may get more/better photographs
    - Individuals with more distinct markings may be more likely to be identified/re-identified when image quality is lower 
    - Cooch and White (2014) describe this as the most ciritical assumption for JS models
6. Sort of? - The study area is confined to locations around Northern Baffin Island (mainly Admiralty Inlet and Eclipse Sound) and Cumberland Sound, but we have not consistently sampled in each location each year 

Prep the environment: 
```{r}
rm(list=ls())

setwd("~/Documents/Master's/Analysis/CMR")

library(RMark)
library(dplyr)
```

Prep the CMR data: 
```{r}
CMR_data_gen <- read.csv("genetic_CMR_data_PRELIM.csv", header = TRUE)

# Inserting zeros insead of NAs in years where there are no sightings: 
CMR_data_gen[is.na(CMR_data_gen)] <- 0

# Remove the id column:
CMR_data_gen <- CMR_data_gen %>% 
  select(-genome_sample_ID) %>% 
  mutate(cmr = NA) %>% 
  relocate(cmr, "X2013")

# Creating a column with 1s and 0s for all years observed: 
for (i in 1:nrow(CMR_data_gen)){
  CMR_data_gen[i,1] <- paste(CMR_data_gen[i,2], CMR_data_gen[i,3], CMR_data_gen[i,4], CMR_data_gen[i,5],
                             CMR_data_gen[i,6], CMR_data_gen[i,7], CMR_data_gen[i,8], CMR_data_gen[i,9], 
                             CMR_data_gen[i,10], CMR_data_gen[i,11],sep = "")}

head(CMR_data_gen)

# Save to .csv:
#write.csv(CMR_data_gen, "CMR_data_genetic.csv")
```

Extract first column with CMR data: 
```{r}
CMR_gen <- CMR_data_gen %>% 
  select(cmr) 

# Change name of first column:
colnames(CMR_gen)[colnames(CMR_gen)=="cmr"] <- "ch"
```

Save the CMR data to a txt file and import as ch data:
```{r, message = FALSE, results = 'hide'} 
write.table(CMR_gen, file = "_genetic_CMR_data.txt", row.names = FALSE, quote = FALSE)

ch_CMR_gen <- import.chdata("_genetic_CMR_data.txt")

summary(ch_CMR_gen)
attach(ch_CMR_gen)
```

Start building model: 
```{r}
kw.proc.gen = process.data(ch_CMR_gen, model = "POPAN")
kw.ddl.gen  = make.design.data(kw.proc.gen)
```

Specify effects to consider on survival and detection probabilities: 
```{r}
# Survival process: 
phi.ct   = list(formula = ~1)      # constant
phi.time = list(formula = ~time)   # year effect 

# Detection (capture) process: 
p.ct   = list(formula = ~1)        # constant
p.time = list(formula = ~time)     # year effect

# Entry process:
pent.ct   = list(formula = ~1)     # constant
pent.time = list(formula = ~time)  # year effect
```

Fit models: 
```{r, message = FALSE, results = 'hide'}
# phi  = survival
# p    = detection (capture)
# pent = entry

# Model 1: constant survival, constant recapture, constant entry 
model.1.gen = mark(kw.proc.gen, kw.ddl.gen, output = FALSE, delete = T,
                   model.parameters = list(Phi = phi.ct, p = p.ct, pent = pent.ct))

# Model 2: time-dependent survival, constant recapture, constant entry
model.2.gen = mark(kw.proc.gen, kw.ddl.gen, output = FALSE, delete = T,
                   model.parameters = list(Phi = phi.time, p = p.ct, pent = pent.ct))

# Model 3: constant survival, time-dependent recapture, constant entry 
model.3.gen = mark(kw.proc.gen, kw.ddl.gen, output = FALSE, delete = T,
                   model.parameters = list(Phi = phi.ct, p = p.time, pent = pent.ct))

# Model 4: constant survival, constant recapture, time-dependent entry 
model.4.gen = mark(kw.proc.gen, kw.ddl.gen, output = FALSE, delete = T,
                   model.parameters = list(Phi = phi.ct, p = p.ct, pent = pent.time))

# Model 5: time-dependent survival, time-dependent recapture, constant entry 
model.5.gen = mark(kw.proc.gen, kw.ddl.gen, output = FALSE, delete = T,
                   model.parameters = list(Phi = phi.time, p = p.time, pent = pent.ct))

# Model 6: time-dependent survival, constant recapture, time-dependent entry 
model.6.gen = mark(kw.proc.gen, kw.ddl.gen, output = FALSE, delete = T,
                   model.parameters = list(Phi = phi.time, p = p.ct, pent = pent.time))

# Model 7: constant survival, time-dependent recapture, time-dependent entry 
model.7.gen = mark(kw.proc.gen, kw.ddl.gen, output = FALSE, delete = T,
                   model.parameters = list(Phi = phi.ct, p = p.time, pent = pent.time))

# Model 8: time-dependent survival, time-dependent recapture, time-dependent entry 
model.8.gen = mark(kw.proc.gen, kw.ddl.gen, output = FALSE, delete = T,
                   model.parameters = list(Phi = phi.time, p = p.time, pent = pent.time))
```

Take a look at AIC values:  
```{r}
AIC_models <- c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5", "Model 6", "Model 7", "Model 8")

AICc_values <- c(summary(model.1.gen)$AICc,
                summary(model.2.gen)$AICc,
                summary(model.3.gen)$AICc,
                summary(model.4.gen)$AICc,
                summary(model.5.gen)$AICc,
                summary(model.6.gen)$AICc,
                summary(model.7.gen)$AICc,
                summary(model.8.gen)$AICc)

AIC_table <- as.data.frame(cbind(AIC_models, AICc_values))
colnames(AIC_table)[1:2] = c("Model", "AICc")

AIC_table$AICc <- as.numeric(AIC_table$AICc)

AIC_table <- AIC_table %>% 
  mutate(delta_AICc = AICc - min(AICc)) 
AIC_table # Model 3 has best support 
```

Model 3 has the best support according to the AIC. Let's take a closer look at the parameter estimates for model 3: 
```{r}
# Estimate for survival (constant):
phi.table = get.real(model.3.gen,"Phi", se = TRUE) # Estimate for survival is 1
phi.table[c("estimate","se","lcl","ucl")][1,]

# Estimate for recapture (time-dependent): 
p.table = get.real(model.3.gen,"p", se= TRUE)
p.table[c("estimate","se","lcl","ucl")][1:9,]

format(p.table, scientific = FALSE)
p.table

# Estimate for entry (constant): 
pent.table = get.real(model.3.gen,"pent", se= TRUE)
pent.table[c("estimate","se","lcl","ucl")][1,]

# Estimate for population size
N.table = get.real(model.3.gen,"N", se= TRUE)
N.table[c("estimate","se","lcl","ucl")][1,]
```

Put estimates for superpopulation size from each model in a table: 
```{r}
N.table1 = get.real(model.1.gen,"N", se= TRUE)
N.table2 = get.real(model.2.gen,"N", se= TRUE)
N.table3 = get.real(model.3.gen,"N", se= TRUE)
N.table4 = get.real(model.4.gen,"N", se= TRUE)
N.table5 = get.real(model.5.gen,"N", se= TRUE)
N.table6 = get.real(model.6.gen,"N", se= TRUE)
N.table7 = get.real(model.7.gen,"N", se= TRUE)
N.table8 = get.real(model.8.gen,"N", se= TRUE)

results_table <- rbind(N.table1[c("estimate","se","lcl","ucl")][1,], 
                       N.table2[c("estimate","se","lcl","ucl")][1,], 
                       N.table3[c("estimate","se","lcl","ucl")][1,], 
                       N.table4[c("estimate","se","lcl","ucl")][1,], 
                       N.table5[c("estimate","se","lcl","ucl")][1,], 
                       N.table6[c("estimate","se","lcl","ucl")][1,], 
                       N.table7[c("estimate","se","lcl","ucl")][1,], 
                       N.table8[c("estimate","se","lcl","ucl")][1,])

rownames(results_table) <- NULL

Model_table <- cbind(AIC_table, results_table)

Model_table <- Model_table %>% 
  arrange(delta_AICc) %>% 
  mutate(adj_estimate = estimate/(2/3),
         adj_se = se/(2/3),
         across(where(is.numeric), ~ round(., 2))) 
Model_table
```

Much smaller estimate than given from the Photo ID CMR model, but probably to be expected given there is a lot less data.

### Pradel Lambda

Start building model: 
```{r}
kw.pradel.proc.gen = process.data(ch_CMR_gen, model = "Pradlambda")
kw.pradel.ddl.gen  = make.design.data(kw.pradel.proc.gen)
```

Specify effects to consider on survival and detection probabilities: 
```{r}
# Survival process: 
phi.pradel.ct  = list(formula = ~1)      # constant
phi.prade.time = list(formula = ~time)   # year effect 

# Detection process: 
p.pradel.ct   = list(formula = ~1)        # constant
p.pradel.time = list(formula = ~time)     # year effect

# We assume that the growth rate is constant
```

Fit models: 
```{r, message = FALSE, results = 'hide'}
# phi  = survival
# p    = detection 
# pent = entry

# Model 1: constant survival, constant recapture
model.pradel.1.gen = mark(kw.pradel.proc.gen, kw.pradel.ddl.gen, 
                          model.parameters = list(Phi = phi.pradel.ct, p = p.pradel.ct))

# Model 2: constant survival, time-dependent recapture
model.pradel.2.gen = mark(kw.pradel.proc.gen, kw.pradel.ddl.gen, 
                          model.parameters = list(Phi = phi.pradel.ct, p = p.pradel.time))

# Model 3: time-dependent survival, constant recapture
model.pradel.3.gen = mark(kw.pradel.proc.gen, kw.pradel.ddl.gen,
                          model.parameters = list(Phi = phi.prade.time, p = p.pradel.ct))

# Model 4: time-dependent survival, time-dependent recapture
model.pradel.4.gen = mark(kw.pradel.proc.gen, kw.pradel.ddl.gen, 
                          model.parameters = list(Phi = phi.prade.time, p = p.pradel.time))

```

AIC values:  
```{r}
AIC_pradel_models_gen <- c("Pradel Model 1", "Pradel Model 2", "Pradel Model 3", "Pradel Model 4")

AICc_pradel_values_gen <- c(summary(model.pradel.1.gen)$AICc,
                            summary(model.pradel.2.gen)$AICc,
                            summary(model.pradel.3.gen)$AICc,
                            summary(model.pradel.4.gen)$AICc)

AIC_pradel_table <- as.data.frame(cbind(AIC_pradel_models_gen, AICc_pradel_values_gen))
colnames(AIC_pradel_table)[1:2] = c("Model", "AICc")

AIC_pradel_table$AICc <- as.numeric(AIC_pradel_table$AICc)

AIC_pradel_table <- AIC_pradel_table %>% 
  mutate(delta_AICc = AICc - min(AICc))
AIC_pradel_table # Model 2 has best support - constant survival, time-dependent recapture
```

Model 2 has the best support according to the AIC. Let's take a closer look at the parameter estimates for model 3: 
```{r}
# Estimate for survival (constant):
phi.pradel.table = get.real(model.pradel.2.gen,"Phi", se = TRUE)
phi.pradel.table[c("estimate","se","lcl","ucl")][1,] # Estimate for survival is 0.83

# Estimate for recapture (time-dependent): 
p.pradel.table = get.real(model.pradel.2.gen,"p", se= TRUE)
p.pradel.table[c("estimate","se","lcl","ucl")][1:11,]

# Estimate for lambda (constant):
lambda.pradel.table = get.real(model.pradel.2.gen,"Lambda", se= TRUE)
lambda.pradel.table[c("estimate","se","lcl","ucl")][1,]

# Get derived estimates: 
model.pradel.2.gen$results$derived

# This didn't really work
```

Put it all in a table: 
```{r}
lamb.table1 = get.real(model.pradel.1.gen,"Lambda", se= TRUE)
lamb.table2 = get.real(model.pradel.2.gen,"Lambda", se= TRUE)
lamb.table3 = get.real(model.pradel.3.gen,"Lambda", se= TRUE)
lamb.table4 = get.real(model.pradel.4.gen,"Lambda", se= TRUE)

results_pradel_table <- rbind(lamb.table1[c("estimate","se","lcl","ucl")][1,], 
                              lamb.table2[c("estimate","se","lcl","ucl")][1,], 
                              lamb.table3[c("estimate","se","lcl","ucl")][1,], 
                              lamb.table4[c("estimate","se","lcl","ucl")][1,] 
                              )

rownames(results_pradel_table) <- NULL

Model_pradel_table <- cbind(AIC_pradel_table, results_pradel_table)

Model_pradel_table <- Model_pradel_table %>% 
  arrange(delta_AICc)  
Model_pradel_table
```

This didn't really work - maybe not enough data?


