---
title: "01.2_PhotoID_CMR_Bayes"
date: "`r Sys.Date()`"
---

Here, we will re-analyze the photo-ID CMR data using a Bayesian framework, as described in Marc and Kéry (2011).

The script was obtained from https://github.com/oliviergimenez/bayes-multistate-jollyseber, modified for these data.

Prep the environment: 
```{r, include = FALSE}
#rm(list=ls())

setwd("~/Documents/Master's/Analysis/CMR")

library(tidyverse)
library(rjags)
library(RMark)
library(R2jags)
library(R2ucare)
```

See 02.2_PhotoID_CMR_Bayes for description of model.

POPAN parameterization of data:
```{r}
popan <- function() {
 
   # Priors and constraints
  for (i in 1:M){
    for (t in 1:(n.occasions-1)){
      phi[i,t] <- mean.phi  # Constant survival 
    } #t
    for (t in 1:n.occasions){
      p[i,t] <- p.time[t]  # Time-dependent capture
    } #t
  } #i
  
  mean.phi ~ dunif(0, 1)  # Prior for mean survival - uniform distribution with min = 0 and max = 1
  #mean.phi ~ dnorm(0.8, 0.3)  # Prior for mean survival - normal distribution with mean = 0.8, sd = 0.3
  psi ~ dunif(0, 1) # Prior for inclusion probability
  
  for(t in 1:n.occasions) {
    p.time[t] ~ dunif(0,1)  # prior for time-dependent capture
  }
  
  # Dirichlet prior for entry probabilities
  for (t in 1:n.occasions){
    beta[t] ~ dgamma(1, 1)  # gamma distribution with shape = 1 and scale = 1
    b[t] <- beta[t] / sum(beta[1:n.occasions])
  }
  
  # Convert entry probs to conditional entry probs
  nu[1] <- b[1]
  for (t in 2:n.occasions){
    nu[t] <- b[t] / (1 - sum(b[1:(t-1)]))
  } #t
  
  # Likelihood
  for (i in 1:M){
    # First occasion
    # State process
    w[i] ~ dbern(psi)  # Draw latent inclusion
    z[i,1] ~ dbern(nu[1])
    # Observation process
    mu1[i] <- z[i,1] * p[i,1] * w[i]
    y[i,1] ~ dbern(mu1[i])
    # Subsequent occasions
    for (t in 2:n.occasions){
      # State process
      q[i,t-1] <- 1 - z[i,t-1]
      mu2[i,t] <- phi[i,t-1] * z[i,t-1] + nu[t] * prod(q[i,1:(t-1)])
      z[i,t] ~ dbern(mu2[i,t])
      # Observation process
      mu3[i,t] <- z[i,t] * p[i,t] * w[i]
      y[i,t] ~ dbern(mu3[i,t])
    } #t
  } #i
  
  # Calculate derived population parameters
  for (i in 1:M){
    for (t in 1:n.occasions){
      u[i,t] <- z[i,t] * w[i]  # Deflated latent state (u)
    }
  }
  for (i in 1:M){
    recruit[i,1] <- u[i,1]
    for (t in 2:n.occasions){
      recruit[i,t] <- (1 - u[i,t-1]) * u[i,t]
    } #t
  } #i
  for (t in 1:n.occasions){
    N[t] <- sum(u[1:M,t])  # Actual population size
    B[t] <- sum(recruit[1:M,t])  # Number of entries
  } #t
  for (i in 1:M){
    Nind[i] <- sum(u[i,1:n.occasions])
    Nalive[i] <- 1 - equals(Nind[i], 0)
  } #i
  for (t in 1:(n.occasions-2)) {
    lambda[t] <- N[t+1]/N[t] # Pradel-Lambda realized population growth rate
    f[t] <- B[t+1]/N[t] # recruitment (per capita entry) rate
  } #t
  Nsuper <- sum(Nalive[]) # Superpopulation size
  mean.lambda <- mean(lambda[]) # mean realized growth rate
}
```

Now let's apply it to our data. 

Ungroup the data:
```{r}
ch_CMR_gen <- import.chdata("_genetic_CMR_data.txt")
popan_ch_gen <- splitCH(ch_CMR_gen$ch)
```

Augment the observed capture histories by nz pseudo-individuals, all with capture histories of 0: 
```{r}
nz <- 200     # Augmenting the data by 200 pseudo-individuals
CH.aug.gen <- rbind(popan_ch_gen, matrix(0, ncol = dim(popan_ch_gen)[2], nrow = nz))
```

Bundle data.
```{r}
bugs.data.gen <- list(y = CH.aug.gen, 
                      n.occasions = dim(CH.aug.gen)[2], 
                      M = dim(CH.aug.gen)[1])

# Would often get an error for "Invalid parent values" - this is  fix for that. 
bugs.data.gen$y[1] <- 1
```

Initial values.
```{r}
zinit <- CH.aug.gen
zinit[zinit==0] <- 1

n.occasions.js <- ncol(CH.aug.gen)

inits <- function(){list(mean.phi = runif(1, 0, 1), 
                         p.time = runif(n.occasions.js, 0, 1), 
                         psi = runif(1, 0, 1), 
                         z = zinit)}
```

Parameters monitored.
```{r}
parameters <- c("psi", "p.time", "mean.phi", "b", "Nsuper", "N", "B", "nu", "lambda", "f", "mean.lambda")
```

MCMC settings.
```{r}
n.iter   <- 10000     # Number of iterations
n.burnin <- 2000      # Number discarded (burn-in)
n.chains <- 3         # Number of chains
```

Call Jags.
```{r}
kw_popan_gen <- R2jags::jags(data  = bugs.data.gen,
                     inits = inits,
                     parameters.to.save = parameters,
                     model.file = popan,
                     n.chains = n.chains,
                     n.iter = n.iter,
                     n.burnin = n.burnin)
kw_popan_gen
```

```{r}
library(mcmcplots)
kw_popan.mcmc <- as.mcmc(kw_popan_gen)

# Traceplot
mcmcplots::traplot(kw_popan.mcmc, parms = "Nsuper", "deviance", "mean.p", "mean.phi")

# Density plot
mcmcplots::denplot(kw_popan.mcmc, parms = "Nsuper", "deviance")
```


