---
title: "01.2_PhotoID_CMR_Bayes"
date: "`r Sys.Date()`"
---

Capture mark recapture of Arctic killer whales using a genetic (whole genome) identification history. Here, we will analyze the genetic CMR data using a Bayesian framework, as described in Marc and Kéry (2011).

The script was obtained from https://github.com/oliviergimenez/bayes-multistate-jollyseber, modified for these data (described in 02.2_PhotoID_CMR_Bayes).

### POPAN Jolly Seber 

Assumptions for Jolly-Seber Mark Recapture models: 
1. Animals retain their tags throughout the experiment 
2. Tags are read properly
3. Sampling is instantaneous
4. Survival probabilities are the same for all animals (marked and unmarked) between each pair of sampling occasions (homogenous survival)
5. Catchability is the same for all animals (marked and unmarked) at each sampling occasion (homogenous catchability)
6. The study area is constant

In the case of killer whales, do we meet the assumptions?
1. Yes - most nicks and scars used for ID are retained through the life of the animal.
2. Yes - we can assume that identified individuals are re-identified reliably. However, there are ways to account for identification error - to be explored later. 
3. Yes - sampling period is short (1 to a few days)
4. Probably? - while survival probabilities might differ between sex and age classes, being "marked" does not affect an individual's survival
5. Unlikely? - Equal catchability could be affected by: 
    - Behaviour: some individuals/groups may be more likely to approach the boat, and thus we may get more/better photographs
    - Individuals with more distinct markings may be more likely to be identified/re-identified when image quality is lower 
    - Cooch and White (2014) describe this as the most ciritical assumption for JS models
6. Sort of? - The study area is confined to locations around Northern Baffin Island (mainly Admiralty Inlet and Eclipse Sound) and Cumberland Sound, but we have not consistently sampled in each location each year 

### Analysis 

Prep the environment: 
```{r, include = FALSE}
setwd("~/Documents/Master's/Analysis/CMR")

library(tidyverse)
library(rjags)
library(R2jags)
library(RMark)
```

See 02.2_PhotoID_CMR_Bayes for description of model.

POPAN parameterization of data:
```{r}
popan <- function() {
 
   # Priors and constraints
  for (i in 1:M){
    for (t in 1:(n.occasions-1)){
      phi[i,t] <- mean.phi  # Constant survival 
    } #t
    for (t in 1:n.occasions){
      p[i,t] <- p.time[t]  # Time-dependent capture
    } #t
  } #i
  
  mean.phi ~ dunif(0, 1)  # Prior for mean survival - uniform distribution with min = 0 and max = 1
  psi ~ dunif(0, 1) # Prior for inclusion probability
  
  for(t in 1:n.occasions) {
    p.time[t] ~ dunif(0,1)  # prior for time-dependent capture
  }
  
  # Dirichlet prior for entry probabilities
  for (t in 1:n.occasions){
    beta[t] ~ dgamma(1, 1)  # gamma distribution with shape = 1 and scale = 1
    b[t] <- beta[t] / sum(beta[1:n.occasions])
  }
  
  # Convert entry probs to conditional entry probs
  nu[1] <- b[1]
  for (t in 2:n.occasions){
    nu[t] <- b[t] / (1 - sum(b[1:(t-1)]))
  } #t
  
  # Likelihood
  for (i in 1:M){
    # First occasion
    # State process
    w[i] ~ dbern(psi)  # Draw latent inclusion
    z[i,1] ~ dbern(nu[1])
    # Observation process
    mu1[i] <- z[i,1] * p[i,1] * w[i]
    y[i,1] ~ dbern(mu1[i])
    # Subsequent occasions
    for (t in 2:n.occasions){
      # State process
      q[i,t-1] <- 1 - z[i,t-1]
      mu2[i,t] <- phi[i,t-1] * z[i,t-1] + nu[t] * prod(q[i,1:(t-1)])
      z[i,t] ~ dbern(mu2[i,t])
      # Observation process
      mu3[i,t] <- z[i,t] * p[i,t] * w[i]
      y[i,t] ~ dbern(mu3[i,t])
    } #t
  } #i
  
  # Calculate derived population parameters
  for (i in 1:M){
    for (t in 1:n.occasions){
      u[i,t] <- z[i,t] * w[i]  # Deflated latent state (u)
    }
  }
  for (i in 1:M){
    recruit[i,1] <- u[i,1]
    for (t in 2:n.occasions){
      recruit[i,t] <- (1 - u[i,t-1]) * u[i,t]
    } #t
  } #i
  for (t in 1:n.occasions){
    N[t] <- sum(u[1:M,t])  # Actual population size
    B[t] <- sum(recruit[1:M,t])  # Number of entries
  } #t
  for (i in 1:M){
    Nind[i] <- sum(u[i,1:n.occasions])
    Nalive[i] <- 1 - equals(Nind[i], 0)
  } #i
  for (t in 1:(n.occasions-1)) {
    lambda[t] <- N[t+1]/N[t] # Lambda realized annual population growth rate
    f[t] <- B[t+1]/N[t] # recruitment (per capita entry) rate
  } #t
  Nsuper <- sum(Nalive[]) # Superpopulation size
  mean.lambda <- (prod(lambda[]))^(1/(n.occasions-1)) # geometric mean realized growth rate 
}
```

Now let's apply it to our data. 

Load and format the genetic capture history data:
```{r}
# Load data and insert zeros instead of NAs in years where there are no sightings: 
CMR_data_gen <- read.csv("genetic_CMR_data_final.csv", header = TRUE) 
CMR_data_gen[is.na(CMR_data_gen)] <- 0

# Remove the id column and add a new column for CH data:
CMR_data_gen <- CMR_data_gen %>% 
  select(-genome_sample_ID) %>% 
  mutate(cmr = NA) %>% 
  relocate(cmr, "X2013")

# Fill the cmr column with 1s and 0s for all years observed: 
for (i in 1:nrow(CMR_data_gen)){
  CMR_data_gen[i,1] <- paste(CMR_data_gen[i,2], CMR_data_gen[i,3], CMR_data_gen[i,4], CMR_data_gen[i,5],
                             CMR_data_gen[i,6], CMR_data_gen[i,7], CMR_data_gen[i,8], CMR_data_gen[i,9], 
                             CMR_data_gen[i,10], CMR_data_gen[i,11],sep = "")}

# Select first column only, rename, and save to txt file
CMR_gen <- CMR_data_gen %>% 
  select(cmr) %>% 
  dplyr::rename(ch = cmr) %>% 
  write.table(file = "_genetic_cmr_data.txt", row.names = FALSE, quote = FALSE)

# Re-import and split data
ch_CMR_gen <- import.chdata("_genetic_cmr_data.txt")
popan_ch_gen <- splitCH(ch_CMR_gen$ch)
```

Augment the observed capture histories by nz pseudo-individuals, all with capture histories of 0: 
```{r}
nz <- 200     # Augmenting the data by 200 pseudo-individuals
CH.aug.gen <- rbind(popan_ch_gen, matrix(0, ncol = dim(popan_ch_gen)[2], nrow = nz))
```

Bundle data.
```{r}
bugs.data.gen <- list(y = CH.aug.gen, 
                      n.occasions = dim(CH.aug.gen)[2], 
                      M = dim(CH.aug.gen)[1])
```

Initial values.
```{r}
zinit <- CH.aug.gen
zinit[zinit==0] <- 1

n.occasions.js <- ncol(CH.aug.gen)

inits <- function(){list(mean.phi = runif(1, 0, 1), 
                         p.time = runif(n.occasions.js, 0, 1), 
                         psi = runif(1, 0, 1), 
                         z = zinit)}
```

Parameters monitored.
```{r}
parameters <- c("psi", "p.time", "mean.phi", "b", "Nsuper", "N", "B", "nu", "lambda", "f", "mean.lambda")
```

MCMC settings.
```{r}
n.iter   <- 50000     # Number of iterations
n.burnin <- 10000     # Number discarded (burn-in)
n.chains <- 3         # Number of chains
```

Call Jags - run model on server.
```{r}
kw_popan_gen <- R2jags::jags(data  = bugs.data.gen,
                     inits = inits,
                     parameters.to.save = parameters,
                     model.file = popan,
                     n.chains = n.chains,
                     n.iter = n.iter,
                     n.burnin = n.burnin)
kw_popan_gen
```

Save run: 
```{r}
#saveRDS(kw_popan_gen, "output/kw_popan_gen_results_sept2024_run1.rds") 

kw_popan_gen <- readRDS("output/kw_popan_gen_results_sept2024_run1.rds")
kw_popan_gen
```


```{r}
library(mcmcplots)
library(coda)

# Convert the results to an mcmc list object
kw_popan_gen.mcmc <- as.mcmc(kw_popan_gen)

# Save mcmc object: 
#saveRDS(kw_popan_gen.mcmc, "output/kw_popan_gen_mcmc_sept2024_run1.rds") 

cols.gen <- c("#66271c", "#b84733", "#dc8374")

# Traceplot
#tiff("plots/kw_popan_gen_traceplot.tiff", units="in", width=8, height=5, res=400)
mcmcplots::traplot(kw_popan_gen.mcmc, parms = c("Nsuper", "mean.lambda", "deviance", "mean.phi"), style = "plain", col = cols.gen)
#dev.off()

# Density plot
#tiff("plots/kw_popan_gen_densityplot.tiff", units="in", width=8, height=5, res=400)
mcmcplots::denplot(kw_popan_gen.mcmc, parms = c("Nsuper", "mean.lambda", "deviance", "mean.phi"), style = "plain", col = cols.gen)
#dev.off()
```

